# The Basic Model of Virus Dynamics

```{r }
#| message: false
#| warning: false
library(deSolve)
library(tidyverse)
```

The basic model of virus dynamics [@nowak2000] tracks three populations: uninfected target cells $x$, infected cells $y$, and free virus particles $v$. Uninfected cells are produced at a constant rate $\lambda$, die at rate $d$, and become infected on contact with virus at rate $\beta$. Infected cells die at rate $a$, and each produces free virus at rate $k$, which is in turn cleared at rate $u$.

$$
\frac{dx}{dt} = \lambda - dx - \beta x v
$$
$$
\frac{dy}{dt} = \beta x v - a y
$$
$$
\frac{dv}{dt} = k y - u v
$$

The basic reproductive ratio of the virus, the average number of newly infected cells arising from one infected cell placed in a fully susceptible cell population, is

$$
R_0 = \frac{\beta \lambda k}{a d u}
$$

The virus can establish an infection only if $R_0 > 1$.

## Set Up the ODE

```{r }
base_ode <- function(time, state, parameters){
	with(as.list(c(state, parameters)),{
		
		dx <- lambda - d*x - beta * x * v
		dy <- beta * x * v - a * y
		dv <- k * y - u * v
		
		return(list(c(dx,dy,dv)))
	})
	
}

t <- seq(0,30,.1)

params <- c(
	lambda =1e5,# Uninfected cell production rate
	d = .1, # Cell Death Rate
	a = .5, # Infected Cell Death Rate
	beta = 2e-7, # "Rate Constant"
	k = 100, # Virus production from infected cell
	u = 5 # Free virus lifespan
)
```

## Solve the System

Guessing initial values from a graph:

```{r }
x0 <- params["lambda"][1]/params["d"][1]
init <- c(x = unname(x0), 
					y = 1, v = 1)

out <- ode(init, t, base_ode, params)

out_df <- as_tibble(as.data.frame(out))
```

Note that the differences in the scales across compartments are reproduced almost immediately, as in the book.

```{r }

compartment_names <- c("Uninfected Cells",
"Infected Cells",
"Free Virus")

p <- out_df %>% 
	setNames(c("time", compartment_names)) %>% 
	gather(compartment, value, -time) %>% 
	mutate(compartment = factor(compartment, compartment_names)) %>% 
	ggplot(aes(time, y = value))+
	geom_line()+
	facet_grid(rows = vars(compartment),scales = "free_y")+
	theme_classic()+
	labs(
		title = "Viral Dynamics of Primary Phase of HIV or SIV Infection",
		x = "Days",
		y = "Count"
	)+
	scale_y_continuous(labels = scales::comma_format(accuracy = 1000))

p	
```

